Slitasje og kjørespor

### 2023-01-31 DRAFT_slitasjeOgKjørespor.Rmd
doc: Spell check and formatting

## About the underlying data
The indicator uses a data set from a standardised field survey of nature types. You can read more about it [here](#naturtyper). I also import a data set called [ANO](#ANO-import)

### Representativity in time and space
The nature type mapping is not random and cannot be said to be area representative. 
The data is from 2018 to present. The data from one field season usually becomes available early the following year.

### Original units
The

@@

<!-- Text here -->

@@

<!-- Text here -->

@@

<!-- Text here -->

@@

<!-- Text here. Note that there are two parallell systems for defining ecosystem properties - one Norwegian and one international (SEEA EA). See man/SEEA EA examples.pdf for help -->

@@

<!-- Text here -->

## Reference state and values

### Reference state

<!-- Text here -->

@@

<!-- Text here -->

@@

<!-- Text here -->

@@

<!-- Text here -->

@@

This indicator uses the dataset
`Naturtyper etter Miljødirektoratets Instruks`, which can be found
[here](https://kartkatalog.geonorge.no/metadata/naturtyper-miljoedirektoratets-instruks/eb48dd19-03da-41e1-afd9-7ebc3079265c).
See [here](#naturtype) for a detailed description of the dataset.

Importing the full dataset can be done like this, but because it can
take a few seconds due to its size, I will cache the data on the P
server.

@@

We also have a separate summary file where the nature types are manually
mapped to the correct NiN-main types. We can use this to find the nature
types associated with the NiN main types that we are interested in.

@@

We are only interested in a few NiN main types and their mapping units
(see [here](https://ninanor.github.io/ecosystemCondition/naturtype.html)
for justification)

@@

Import back in

```{r}
naturetypes <- readRDS(paste0(pData, "rds/naturtyper.rds"))
```

@@

Keep only the relevant variables.

@@

Names starting with a number is problematic, so adding a prefix

@@

Now I need to create a single row per locality with a new variable which
is a product of the four variables "7SE", "7TK", "PRTK" and "PRSL". I
will create a new data table where I calculate the new variable which I
then paste back into the sf object.

@@

7TK and 7SE use a 4 step scale ![Four step condition
scale](images/ninAR4b.PNG)

I want to base this indicator on whichever of the variables have the
highest value (worst condition). This table shows that a locality may
have been assessed with both a PR variable and a NiN variable

@@

```{r}
head(naturetypes_wide)
anyDuplicated(naturetypes_wide$identifikasjon_lokalid)
```

@@

When both variables are NA, we get -Inf. There were nine cases of this.
Removing these now.

@@

This variable is also recorded in GRUK. The nature type dataset I'm
working on here includes this data already (presently only 2021
included). GRUK also records a related variable:
`% cover in 5m radii circles`. This data is not published. In any case
it is better to use the harmonized dataset.

@@

@@

This dataset only contains data from year 2019 and 2021. We need to
update this dataset later.

@@

@@

We want to use area weighting in this indicator, so we can use this
percentage cover data to calculate the area. Note that both datasets use
m2 as area units.

@@

Let's now look at the distribution of the variables. First I need to
seperate the variable name fro the values themselves. Now the data looks
like this:

@@

@@

We need to combine the nature type dataset with the ANO dataset. I will
add a column `origin` to show where the data comes from. I will also and
add data about the main ecosystem.

@@

I use `dplyr::select` to can reduce the number of columns to keep things
a bit more tidy.

@@

The ANO data is point data, and the nature type data is multipolygon.
Because later we want to save this as a shape file, we cannot have a
mixed class type. I will therefore convert the points to polygons. I use
the area column to calculate a radii that gives that area.

@@

The area calculation seems to have worked fine.

@@

The new data set contains **only polygons**.

Grouping and tallying the data for plotting

```{r}
temp <- as.data.frame(table(slitasje_data$slitasje))
```

#### Distribution of Slitasje scores
```{r, fig.cap="Slitasje scores (ANO Naturetype (and GRUK) data combined). The score for a locality equals the highest (worst) score of the related variables 7TK, 7SE, PRSL and PRTK."}
ggplot(temp, aes(x = Var1,
                 y = Freq))+
  geom_bar(stat="identity",
           fill="grey",
           colour = "black")+
  theme_bw(base_size = 12)+
  labs(x = "slitasje score",
       y = "Number of localities")
```

@@

So the ANO data is not very important here, but it can become more
important in the future.

@@

The GIS layers are used to crop out marine areas, and to define accounting areas, respetively.

@@

I can scale the indicator for each polygon, or I can chose to aggregate
them first. If the scaled value is representative and precise at the
polygon level, then I could scale at that level. However, the combined
surveyed area is still so little of the total area of Norway, so that
only producing indicator values for these leaves the indicator without
much value for regional assessments. Later when we produce assessments
and indicator values at the regional level, we cannot simply do an area
weighting of the polygons in each region. This is because we don't want
to assume that the polygons are representative far outside of the mapped
area. But perhaps we can assume them to be representative inside
*homogeneous ecological areas*. That's where the infrastructure index
comes in. Here's the plan:

@@

2.  take the simplified infrastructure index and
    extract the indicator value that intersect with a given cell, and
    extrapolate the area weighted mean of those values to all similar cell within a
    given region (`ea_spread()`). Errors are calculated from bootstrapping.

3. Take the new area weighted indicator values for the homogeneous impact area (HIA) and match them with ecosystem occurrences of the same HIA. The errors are the same as for the HIAs.

3.  calculate an indicator value for a region (accounting area) by doing a weighted
    average based of the relative area of ecosystem occurrences. Errors shold be carried somehow, perhaps via weighted resampling

#### Scaled variable

The variable is ordinal, not continuous. See fig. \@ref(fig:eight-step).
I could create another variable which was more like original range
scale. Although not perfect, I will use the median value for each
category.

@@

@@

@@

I want to use the [Homogeneous Impact Areas](#HIA) (HIA) to define
smaller regions into which I can extrapolate the indicator values. This
data is generated by discretizing the Norwegian Infrastructure Index. I
refer to the ordinal values of the four impact area classes as their
*Human Impact Factor* (HIF).

@@

I want to check that HIF is in fact a good predictor for _slitasje_.

@@

```{r, echo=F}
# silently read cached file

HIA_regions <- readRDS(paste0(pData, "cache/HIA_regions.rds"))
```

@@

#### Validate
I now have 20 unique areas (for each main ecosystem) that I will, given there is data, extrapolate
indicator values over

###### Subset ETs
I will subset the `sliatsje_data` in the three ecosystems. 
Note that only for open wetlands do we have good ecosystem delineation maps to base a spatial averaging on.

@@

@@

This rather dense figure needs to be looked at carefully. 
In the top row we see that the sample size (number of polygons) varies between ecosystems. For naturally open ecosystems many polygons are from high infrastructure areas, not surprisingly, and there are few points for the lowest impact class (zero such polygins in _Sørlandet_ for example). We there may not be able to extrapolate indicator values in areas in _Sørlandet_ with low levels of infrastructure. We can also from the middle row see the association between the indicator values (area weighted means) and the modified infrastructure index (HIF). We should then not put too much weight on data were there is little data (low n) and/or high standard error (sd, bottom row). For example, with Naturally open ecosystems, Midt-Norge has the overall lowest errors, and there we also see the kind of relationship between the indicator and the pressure that we want to see (middle row). The pattern is repeated for all regions except Østlandet, but there we also see that HIF class 3 has very little data, and a high sd, and this could explain the discrepancy (but it doesn't support the general relationship either).

@@

```{r, eval=F}
corrCheck <- st_intersection(slitasje_data, HIA_regions)
saveRDS(corrCheck, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/corrCheck.rds")
```
```{r, include=F}
#Silently read cached data
corrCheck <- readRDS(paste0(pData, "cache/corrCheck.rds"))
```

@@

When down weighting the importance of those groups/categories that have very little The figure above I think supports the indicator-pressure relationship and justifies using the HIA+region intersections as local reference areas (i.e. homogeneous impact areas).

We can also calculate Kendall tau b coefficients, although I'm not complete sure this is valid when we have ordinal data with only four levels, since there will be a lot of ties.

@@

From the figures above it appears that the relationship is quite week in general. However, the sheer number of data points suggest that group differences will be very reliable.
_Naturlig åpne områder_ has the lowest correlation between the indicator and the homogeneous area classes, and of the five regions, _Nord-Norge_ has the lowest correlation.

#### Looking at the HIA
Here is a view of the data zooming in on Trondheim

@@

natOpen_slitasje_extr <- ea_spread(indicator_data = natOpen,
                         indicator = i_2020,
                         regions = HIA_regions,
                         groups = region_HIF,
                         threshold = 100)
#saveRDS(wetlands_slitasje_extr, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/wetlands_slitasje_extr.rds")
#saveRDS(seminat_slitasje_extr, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/seminat_slitasje_extr.rds")
#saveRDS(natOpen_slitasje_extr, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/natOpen_slitasje_extr.rds")
```
```{r, include=F}

wetlands_slitasje_extr <- readRDS(paste0(pData, "cache/wetlands_slitasje_extr.rds"))
seminat_slitasje_extr  <- readRDS(paste0(pData, "cache/seminat_slitasje_extr.rds"))
natOpen_slitasje_extr  <- readRDS(paste0(pData, "cache/natOpen_slitasje_extr.rds"))
```

@@

@@

There is some, but very little variation both between the ecosystems and the four classes in the modified infrastructure index. Note however, that for wetland and semi-natural areas we did not extrapolate values for the most urban areas, and that for naturally open areas we did not extrapolate to the least affected areas.

_These maps are easy to misunderstand, and they should therefor be communicated with caution._

Now, there are multiple way to go proceed. We have indicator maps (vector data), and we have HIA area, and we have the HIA with indicator data (also vector data).

@@

1. crop the extrapolated indicator maps to ecosystem delineation maps, and then;

@@

Step 1 above requires an st_interactions on two very large vactor data sets. This should be avoided.

@@

1. rasterize the extrapolated indicator map, using the ET map as a tamplate, and

@@

1. Draw n random point samples from the ET map, and

@@

This is perhaps even faster. I will try the last option, even though it will not provide spatailly explicit indicator values for the entire ecosystem delineation, like option 2 does. Perhaps I will revisit option 2 later.

### Ecosystem map
Importing the wetland map.

This is a quite heavy operation and should be done on a server.
```{r, eval=F}
mireET <- sf::st_read("/data/R/Kladd/Myr/Myrmodell.gdb")
```

This doesn't take that long.
```{r, eval=F}
mireET <- st_transform(mireET, 25833)
```

This take about 10 min. Not sure I need it.
```{r, eval=F}
mireET <- st_make_valid(mireET)
```

This takes a long!! time, and should be AVOIDED. 
```{r, eval=F}
wetlands_slitasje_extr_cropped <- sf::st_intersection(wetlands_slitasje_extr, mireET)
```

Instead I will draw some random point samples.
```{r}
wetland_points <- st_sample(mireET, 10000)
```

And intersect this with the extrapolated indicator values.
```{r}
wetland_points_match <- st_intersection(wetlands_slitasje_extr, wetland_points)
```

Create an empty grid, 50 x 50 m resolution
```{r}
grd = st_as_stars(st_bbox(wetlands_slitasje_extr), dx = 50, dy = 50, values = NA_real_)
```

Rasterize the indicator
```{r}
wetlands_slitasje_extr_rast <- st_rasterize(wetlands_slitasje_extr, grd)
```

And extract the values
```{r}
wetland_points_extracted <- st_extract(wetlands_slitasje_extr, wetland_points)
```

@@

```{r, eval=F}
wetlands_slitasje_extr_cropped_reg <- ea_spread(indicator_data = wetlands_slitasje_extr_cropped,
                                      indicator = i_2020,
                                      regions = regions,
                                      groups = ?)
```

@@

According to the workflow specifications, I will export the indicator
map at its highest resolution, for use in various web interfaces etc.

But I also need to export a map of the indicator where we extrapolate
the indicator values to the homogeneous areas. This map should be
interpreted with caution, since they seem to provide explicit indicator
values for areas where there is no data. This map is mainly for
aggregation purposes. I will mark this map with the prefix
*homogeneous-areas*.

I will also export a map of the *homogeneous areas*, over which I will
aggregate the indicator values. This map might be useful in later
indicator work.

For the finest resolution indicator map I will not include columns for
uncertainty (because this will only be estimated at the aggregated
level), or reference levels (because there is no spatial variation).

### Sliatsje in wetlands

@@

```{r}
exp_wetland <- select(slitasje_data[slitasje_data$hovedøkosystem == "Våtmark",],
              origin,
              area,
              i_2020,
              SHAPE
              )
```

@@

```{r, eval=F}
st_write(exp_wetland, "indicators/indicatorMaps/slitasje/wetland_slitasje.shp",
         append = F)
```

### Sliatsje in wetlands, extrapolated to homegeneous impact areas.

>>>>

## About the underlying data

The indicator uses a data set from a standardised field survey of nature
types. You can read more about the preliminary analyses
[here](#naturtype). See also the [official
site](https://www.miljodirektoratet.no/ansvarsomrader/overvaking-arealplanlegging/naturkartlegging/naturtyper/)
of the Environmental Agency. I also import a data set called
[ANO](#ANO-import), which you can read about
[here](https://www.miljodirektoratet.no/ansvarsomrader/overvaking-arealplanlegging/miljoovervaking/overvakingsprogrammer/natur-pa-land/arealrepresentativ-naturovervakning-ano/)

### Representativity in time and space

The nature type mapping is not random and cannot be said to be area
representative. The ANO dataset however, is area representative. The
data is from 2018 to present. The data from one field season usually
becomes available early the following year.

### Original units

The variables are recorded on a coarse four-sped ordinal scale (Fig.
\@ref(fig:four-step)).

@@

The data goes back to 2018, but is because of data shortage, it is
bulked into one time step. I then use the mean date for the raw data,
and define the variable as belonging to the year 2020 (read more
[here](#scaled-slitasje-variable)).

@@

For a run through of the data set, see [here](#naturtype).

@@

The indicator is tagged to the *Økologisk egenskap* called
**Primærproduksjon** (Primary productivity). This is tentative, and
perhaps *abiotiske forhold* is more suited. But the thought behind the
chiose is that *sliatsje* affect the potential for primary productivity.

@@

The indicator is tagged as a **Physical state characteristics**
indicator. This is quite clear. It's a tpe of abiotic characteritic that
has to do with the physical structure rather than the chemical
composition.

@@

Slitasje is not thought to be very collinear with any other indicator at
the present.

## Reference condition and values

### Reference condition

The reference condition is one with minimal negative human impact. This
is also true for semi-natural ecosystems. For example, in the reference
condition, *slitasje* in semi-natural ecosystems is defined to be zero,
even though at no point in time would this condition be realized.

@@

The upper reference value is 0 (no *slitasje*), and the lower reference
value is 100%. Note that 100% slitasje doesn't mean that all the area
must be scared or damaged, but that all hypothetical quadrats around a
point is affected (the variable is frequency-driven, and not
amount-driven). The threshold for good ecosystem condition is 10%
damage. This is on hte conservative side. Perhaps is should be set
somewhat higher, like 20 or 30%.

Read about the normalisation [here](##scaled-slitasje-variable).

@@

Uncertaintiesare estimated for aggregated indicator values by
bootstrapping individual indicator values 1000 times and calculating a
distribution of area weighted means. When aggregating a second time,
from homogeneous impact areas to accounting areas, we assume a normal
distribution around the indicator values, with the already mentioned
errors, and sample n times from these and combine the resamples into an
a new, area weighted, distribution. The errors for the accounting areas
thus represents both the spatial variation, and the precision, of the
indicator values within the accounting areas.

@@

No additional references.

@@

This indicator uses the data set
`Naturtyper etter Miljødirektoratets Instruks`, which can be found
[here](https://kartkatalog.geonorge.no/metadata/naturtyper-miljoedirektoratets-instruks/eb48dd19-03da-41e1-afd9-7ebc3079265c).
See also [here](#naturtype) for a detailed description of the data set.

<!-- Importing the full data set can be done like this, but because it can -->

<!-- take a few seconds due to its size, I will cache the data on the P -->

<!-- server after subsetting. -->

@@

We also have a separate [summary file](##exp-natureType-summary) where
the nature types are manually mapped to the correct NiN-main types. We
can use this to find the nature types associated with the NiN main types
that we are interested in.

@@

We are only interested in a few NiN main types and their mapping units
(see [here](#naturtype) for justification)

@@

```{r, include=F}
# Import back in
naturetypes <- readRDS(paste0(pData, "rds/naturtyper.rds"))
```

@@

Keep only the four relevant variables.

@@

Column names starting with a number is problematic, so adding a prefix

@@

Now I need to create a single row per locality with a new variable which
is a product/function of the four variables "7SE", "7TK", "PRTK" and
"PRSL". I will create a new data table where I calculate the new
variable which I then paste back into the sf object.

@@

7TK and 7SE use a 4 step scale

```{r four-step, fig.cap="Four step condition scale"}
knitr::include_graphics("images/ninAR4b.PNG")
```

I want to base this indicator on whichever of the variables have the
highest value (worst condition). This table shows that a locality may
have been assessed with both a PR variable and a NiN variable:

@@

```{r}
head(naturetypes_wide)
```

```{r}
anyDuplicated(naturetypes_wide$identifikasjon_lokalid)
```

@@

When both variables are NA, we get -Inf. There were nine cases of this
(in 2023). Removing these now.

@@

This variable is also recorded in
[GRUK](https://www.nina.no/Naturmangfold/Trua-natur/%C3%85pen-grunnlendt-kalkmark).
The nature type data set I'm working on here includes this data already
(presently only 2021 included). GRUK also records a related variable:
`% cover in 5m radii circles`, which si much more detailed. This data is
not published. In any case it is better to use the harmonized data set
in our case.

@@

> *ANO could harmonie better with the Naturetype data set if it included
> 7SE and 7TK for Naturally open and Semi natural ecosystem.*

@@

This data set only contains data from year 2019 and 2021. We need to
update this data set later. It is not clear why there is no data from
2020.

@@

> *All of 2019 ANO data is excluded because of missing information*

@@

We want to use area weighting in this indicator, so we can use this
percentage cover data to calculate the area. Note that both data sets
use m^2^ as area units.

@@

Let's now look at the distribution of the variables. First I need to
separate the variable name from the values themselves. Now the data
looks like this:

@@

> *The ANO field app should not allow user to leave blank cells. This
> resulted in the exclusion of data.*

@@

We need to combine the nature type data set with the ANO data set. I
will add a column `origin` to show where the data comes from. I will
also add a column with the main ecosystem.

@@

I use `dplyr::select` to reduce the number of columns to keep things a
bit more tidy.

@@

The ANO data is point data, and the nature type data is *multipolygon*.
Because later we might want to save this as a shape file, we cannot have
a mixed class type. I will therefore convert the points to polygons. I
use the area column to calculate a radii that gives that area.

@@

The area calculation seems to have worked fine. Checking that the new
data set contains **only polygons**.

@@

Ok.

#### Distribution of Slitasje scores

```{r, fig.cap="Slitasje scores (ANO Naturetype (and GRUK) data combined). The score for a locality equals the highest (worst) score of the related variables 7TK, 7SE, PRSL and PRTK."}
temp <- as.data.frame(table(slitasje_data$slitasje))
ggplot(temp, aes(x = Var1,
                 y = Freq))+
  geom_bar(stat="identity",
           fill="grey",
           colour = "black")+
  theme_bw(base_size = 12)+
  labs(x = "slitasje score",
       y = "Number of localities")
```

@@

So the ANO data is not very important here, but it can become more
important in the future so good to have them included in the workflow.

@@

The GIS layers are used to crop out marine areas, and to define
accounting areas, respectively.

@@

I can scale the indicator for each polygon, or I can chose to aggregate
them first. If the scaled value is representative and precise at the
polygon level, then I could scale at that level. I think they are.

However, the combined surveyed area is a very small fraction of the
total area of Norway, so that only producing indicator values for the
mapped areas leaves the indicator without much value for regional
assessments. When we do regional assessments and calcualte regional
indicator values, we cannot simply do an area weighting of the polygons
in each region. This is because we don't want to assume that the
polygons are representative far outside of the mapped area. But perhaps
we can assume them to be representative inside *homogeneous ecological
areas,* or *Homoegeneous Impact Areas* (HIAs). That's where the
infrastructure index comes in. Here's the plan:

@@

2.  take a simplified infrastructure index (vector data set with HIAs)
    and extract the corresponding indicator values that intersect with a
    given HIA, and extrapolate the area weighted mean of those values to
    the entire HIA withing a given geographical region. Errors are
    calculated from bootstrapping.

3.  Take the new area weighted indicator values for the HIAs and match
    them with ecosystem occurrences of the same HIA (possibly using
    randomly sampled points from the available ecosystem occurrences).
    The errors are the same as for the HIAs.

4.  calculate an indicator value for a region (accounting area) by doing
    a weighted average based of the relative area of ecosystem
    occurrences. Errors should be carried somehow, perhaps via weighted
    resampling.

#### Scaled variable {#scaled-slitasje-variable}

The variable is ordinal, not continuous. See fig. \@ref(fig:eight-step).
I could create another variable which was more like original range
scale. Although not perfect, I will use the median value for each
category. I.e. when the variable is recorded as 2 (meaning from 1/16th
to 1/2), this is converted to 31.25%, i.e. in the middle between the two
end points of the category.

@@

Also, the threshold value is set quite conservativly, and should be
discussed.

@@

Assigning the indicator to year 2020.

@@

I want to use the [Homogeneous Impact Areas](#HIA) (HIA) to define
smaller regions into which I can extrapolate the indicator values. This
data is generated by discretizing the Norwegian Infrastructure Index. I
refer to the ordinal values of the four HIA classes as their *Human
Impact Factor* (HIF). This is just to keep the approach separate from
the Norwegian Infrastructure Index.

@@

I want to check that HIF is in fact a good predictor for *slitasje*.

@@

```{r, include=F}
# silently read cached file
HIA_regions <- readRDS(paste0(pData, "cache/HIA_regions.rds"))
```

Create a new column by crossing region and HIF

@@

#### Validate

I now have 20 unique areas (for each main ecosystem) that I will, given
there is data, extrapolate indicator values over.

###### Subset ETs

I will subset the `sliatsje_data` in the three ecosystems. Note that
only for open wetland do we have good ecosystem delineation maps to base
a spatial averaging on.

@@

Creating some summary statistics.

@@

This rather dense figure needs to be looked at carefully. In the top row
we see that the sample size (number of polygons) varies between
ecosystems. For naturally open ecosystems many polygons are from high
infrastructure areas, not surprisingly, and there are few points for the
lowest impact class (zero such polygons in *Sørlandet* for example). We
may therefore not be able to extrapolate indicator values for Naturally
open areas in in *Sørlandet* with low levels of infrastructure. We can
also from the middle row see the association between the indicator
values (area weighted means) and the modified infrastructure index
(HIF). We should then not put too much weight on the bars were there is
little data (low n) and/or high standard error (sd, bottom row). For
example, with Naturally open ecosystems, Midt-Norge has the overall
lowest errors, and there we also see the kind of relationship between
the indicator and the pressure that we want to see (middle row). The
pattern is repeated for all regions except Østlandet, but there we also
see that HIF class 3 has very little data, and a high sd, and this could
explain the discrepancy (but it doesn't support the general relationship
either).

@@

```{r, eval=F}
corrCheck <- st_intersection(slitasje_data, HIA_regions)
saveRDS(corrCheck, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/corrCheck.rds")
```

```{r, include=F}
#Silently read cached data
corrCheck <- readRDS(paste0(pData, "cache/corrCheck.rds"))
```

@@

The figure above I think supports the indicator-pressure relationship
and justifies using the HIA+region intersections as local reference
areas. Especially when down-weighting the importance of those
groups/categories that have very little data.

We can also calculate *Kendall tau b* coefficients, although I'm not
complete sure this is valid when we have ordinal data with only four
levels, since there will be a lot of ties.

@@

From the figures above it appears that the relationship is quite week in
general. However, the sheer number of data points suggest that group
differences will be very reliable. *Naturlig åpne områder* has the
lowest correlation between the indicator and the homogeneous area
classes, and of the five regions, *Nord-Norge* has the lowest
correlation.

#### Looking at the HIA

Here is a view of the data zooming in on Trondheim

@@

natOpen_slitasje_extr <- ea_spread(indicator_data = natOpen,
                         indicator = i_2020,
                         regions = HIA_regions,
                         groups = region_HIF,
                         threshold = 100)
#saveRDS(wetlands_slitasje_extr, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/wetlands_slitasje_extr.rds")
#saveRDS(seminat_slitasje_extr, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/seminat_slitasje_extr.rds")
#saveRDS(natOpen_slitasje_extr, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/natOpen_slitasje_extr.rds")
```

```{r, include=F}
wetlands_slitasje_extr <- readRDS(paste0(pData, "cache/wetlands_slitasje_extr.rds"))
seminat_slitasje_extr  <- readRDS(paste0(pData, "cache/seminat_slitasje_extr.rds"))
natOpen_slitasje_extr  <- readRDS(paste0(pData, "cache/natOpen_slitasje_extr.rds"))
```

@@

Cropping the data to the bbox

@@

There is some, but very little variation, both between the ecosystems
and between the four HIF classes in the modified infrastructure index.
Note however, that for wetland and semi-natural areas we did not
extrapolate values for the most urban areas, and that for naturally open
areas we did not extrapolate to the least affected areas.

> *The maps abov are easy to misunderstand, and they should therefor be
> communicated with caution. They should be considered as temporary
> files or as by-products.*

Now, there are multiple way to proceed. We have

-   indicator maps (vector data), and we have

-   total HIA areas, and we have the

-   extrapolated indicator maps (HIA maps with aggregated indicator
    values and errors, also vector data).

-   We also have, for wetlands, an ecosystem type (ET) map.

@@

1.  crop (i.e. intersect) the extrapolated indicator maps to ecosystem
    delineation maps, and then;

@@

Step 1 above requires an st_interactions on two very large vector data
sets with millions of polygons. This should be avoided.

@@

1.  rasterize the extrapolated indicator map, using the ET map as a
    template, and

@@

1.  Draw *n* random point samples from the ET map, and

@@

This is perhaps even faster again. I will try the last option, even
though it will not provide spatially explicit indicator values for the
entire ecosystem delineation, like option 2 does. Perhaps I will revisit
option 2 later.

### Ecosystem map

Importing the wetland map.

This is a quite heavy operation and should be done on a server.

```{r, eval=F}
mireET <- sf::st_read("/data/R/Kladd/Myr/Myrmodell.gdb")
```

This doesn't take that long.

```{r, eval=F}
mireET <- st_transform(mireET, 25833)
```

This take about 10 min. Not sure I need it.

```{r, eval=F}
mireET <- st_make_valid(mireET)
```

This takes a long!! time, and should be AVOIDED.

```{r, eval=F}
wetlands_slitasje_extr_cropped <- sf::st_intersection(wetlands_slitasje_extr, mireET)
```

Instead I will draw some random point samples.

```{r, eval=F}
wetland_points <- st_sample(mireET, 10000)
```

And intersect this with the extrapolated indicator values.

```{r, eval=F}
wetland_points_match <- st_intersection(wetlands_slitasje_extr, wetland_points)
```

And extract the values

```{r, eval=F}
wetland_points_extracted <- st_intersection(wetlands_slitasje_extr, wetland_points)
```

If we went for option 2, we could instead do something like this, and

create an empty grid over Norway, 50 x 50 m resolution

<!--# This we should perhaps do once, in a separate misc chapter, and make available for all chapters -->

```{r, eval=F}
grd = st_as_stars(st_bbox(wetlands_slitasje_extr), dx = 50, dy = 50, values = NA_real_)
```

and then rasterize the indicator

```{r, eval=F}
wetlands_slitasje_extr_rast <- st_rasterize(wetlands_slitasje_extr, grd)
```

To get a perfect alignment with the ET map, we should change `grd` with
the ET raster map.

Then we can use starts filtering with `[ .. ]` to mask out the areas
that do not belong to the ET.

@@

We need to aggreate the ET occurence and their indicator values, to a
regional delineation. This functionality will be developed in `eaTools`.
We need to produce distributions for the indicator valeus for each ET
occurrence (or groups of them) and resample the distributions wiht an
area weighting.

@@

According to the workflow specifications, I will export the indicator
map at its highest resolution, for use in various web interfaces etc.
For the finest resolution indicator map I will not include columns for
uncertainty (because this will only be estimated at the aggregated
level), or reference levels (because there is no spatial variation).

But I will also export a map of the indicator where we extrapolate the
indicator values to the homogeneous areas. This map should be
interpreted with caution, since they seem to provide explicit indicator
values for areas where there is no data. This map is mainly for
aggregation purposes. I will mark this map with the prefix
*homogeneous-areas*. I can also export this same mappen after masking it
with the ET map, and this output must come with the same discalimers.

Finally I will export a map of the five accounting areas with
wall-to-wall indicator value and errors. This map can also contain data
covering other ETs. Otherwise the maps would not we visually
interpretative.

### Slitasje in wetlands

@@

```{r, eval=F}
# keep col names short and standardized
exp_wetland <- ...
```

```{r, eval=F}
exp_wetland_HIAextrapolated <- …
```

```{r, eval=F}
exp_wetland_AAextrapolated <- …
```

@@

```{r, eval=F}
st_write(exp_wetland, "indicators/indicatorMaps/slitasje/wetland_slitasje.shp",
         append = F)

st_write(exp_wetlandHIAextrapolated, 
         "indicators/indicatorMaps/extrapolated_to_homogeneous_areas/slitasje/homogeneous_areas_wetland_slitasje.shp",
         append = F)

st_write(exp_wetland, "indicators/indicatorMaps/extrapolated_to_accounting_areas/slitasje/accounting_areas_wetland_slitasje.shp",
         append = F)
```

### 2023-01-31 naturtype.Rmd
doc: Spell check and formatting

# Nature types - Part 1 {#naturtype1}

**On the application of the naturetype dataset - Part 1 - Data exploration and cleaning**

@@

```{r, include=FALSE}
library(knitr)
library(sf)
library(tmap)
library(ggplot2)
library(data.table)
library(stringr)
library(units)
library(dplyr)
knitr::opts_chunk$set(echo = TRUE)
```

@@

I will start by excluding all but the three target ecosystems.
```{r}
dat <- dat[dat$hovedøkosystem %in% target,]
```

@@

  group_by(NiN_mainType, NiN_mainTypeCoverage) %>%
  mutate(km2 = sum(km2)) %>%
  slice_head()%>%
  select(Ecosystem,
         NiN_mainType,
         NiN_mainTypeCoverage,
         km2)%>%
  ungroup() %>%
  mutate(NiN_mainType = fct_reorder(NiN_mainType, km2)) %>%
  ggplot(aes(x = NiN_mainType,
             y = km2,
             fill = NiN_mainTypeCoverage))+
    geom_bar(stat = "identity",
             colour = "grey40",
             position = "stack")+
    theme_bw(base_size = mySize)+
    coord_flip()+
  labs(x = "NiN main type",
       y = expression(km^2),
       fill = "Spatial coverage")+
  scale_fill_brewer(palette = "Set1")+
  facet_wrap(vars(Ecosystem),
             scales = "free",
             labeller = label_wrap_gen(width=25))

@@

## Export data
Exporting the table with the summary statistics and other information that I compled for each nature type in the three main ecosystems. This can be merged with the main dataset in order to subset according to a specific research question.

@@

# Appendix 1
Here is a deeper look into the causes for why some localities are recorded with seemingly erroneous NiN mapping units (and hence NiN main types)

>>>>

# Nature types {#naturtype}

**On the application of the nature type dataset**

*Data exploration and an analyses of thematic coverage*

@@

```{r, include=FALSE}
library(knitr)
library(sf)
library(tmap)
library(tidyr)
library(ggplot2)
library(data.table)
library(stringr)
library(forcats)
library(RColorBrewer)
library(units)
library(egg)
library(dplyr)
knitr::opts_chunk$set(echo = TRUE)
```

@@

I will start by excluding all but the three target ecosystems.
```{r}
target <- c("Semi-naturlig mark","Våtmark","Naturlig åpne områder under skoggrensa")
dat <- dat[dat$hovedøkosystem %in% target,]
```

@@

  group_by(NiN_mainType, NiN_mainTypeCoverage) %>%
  mutate(km2 = sum(km2)) %>%
  slice_head()%>%
  select(Ecosystem,
         NiN_mainType,
         NiN_mainTypeCoverage,
         km2)%>%
  ungroup() %>%
  mutate(NiN_mainType = forcats::fct_reorder(NiN_mainType, km2)) %>%
  ggplot(aes(x = NiN_mainType,
             y = km2,
             fill = NiN_mainTypeCoverage))+
    geom_bar(stat = "identity",
             colour = "grey40",
             position = "stack")+
    theme_bw(base_size = mySize)+
    coord_flip()+
  labs(x = "NiN main type",
       y = expression(km^2),
       fill = "Spatial coverage")+
  scale_fill_brewer(palette = "Set1")+
  facet_wrap(vars(Ecosystem),
             scales = "free",
             labeller = label_wrap_gen(width=25))

@@

## Export data {#exp-natureType-summary}
Exporting the table with the summary statistics and other information that I compled for each nature type in the three main ecosystems. This can be merged with the main dataset in order to subset according to a specific research question.

@@

## Appendix 1
Here is a deeper look into the causes for why some localities are recorded with seemingly erroneous NiN mapping units (and hence NiN main types)
